---
title: "Data Provenance and Import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Provenance and Import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`qiblalab` ships a cleaned version of Dan Gibson's *Early Islamic Qibla Database*
(2021), published under CC BY 4.0 at
<https://doi.org/10.6084/m9.figshare.13570655.v2>.
This vignette explains:

- what the shipped dataset contains and how it was cleaned,
- how to download a fresh copy directly from Figshare,
- the three-destination reference table, and
- the data quality validation step.

## The shipped dataset

```{r load}
library(qiblalab)

data(gibson_qibla)
data(destinations)

dplyr::glimpse(gibson_qibla)
```

The dataset has `r nrow(gibson_qibla)` rows (one per mosque) and
`r ncol(gibson_qibla)` columns. The key analytical columns are:

| Column | Description |
|---|---|
| `latitude`, `longitude` | WGS 84 coordinates (decimal degrees) |
| `azimuth` | Observed orientation of the qibla wall (degrees, 0–360) |
| `year_ce_min`, `year_ce_max` | Estimated construction date range (CE) |
| `gibson_classification` | Gibson's published destination classification |

Columns with `NA` carry genuine uncertainty; they are never silently imputed.

### Provenance attributes

The dataset carries a `provenance` attribute recording the source DOI and
download metadata, plus an `audit_log` of every cleaning decision.

```{r provenance}
prov <- attr(gibson_qibla, "provenance")
cat("Source DOI  :", prov$doi,    "\n")
cat("Access date :", prov$access_date, "\n")

# First three audit-log entries
head(attr(gibson_qibla, "audit_log"), 3)
```

### Azimuth coverage

Not all mosques have a measured azimuth. The analysis functions handle
`NA` azimuths transparently — they are excluded from calculations and
flagged in output.

```{r az-coverage}
has_az <- !is.na(gibson_qibla$azimuth)
cat(sprintf("%d / %d mosques have a measured azimuth (%.0f%%).\n",
            sum(has_az), nrow(gibson_qibla),
            100 * mean(has_az)))
```

### Date coverage

```{r date-coverage}
has_min <- !is.na(gibson_qibla$year_ce_min)
has_max <- !is.na(gibson_qibla$year_ce_max)
cat(sprintf("year_ce_min: %d mosques  |  year_ce_max: %d mosques\n",
            sum(has_min), sum(has_max)))
```

## The destinations table

```{r destinations}
destinations
```

Three candidate destinations are shipped. Hypotheses are specified
against these identifiers; the package computes great-circle bearings
from each mosque to each destination on demand.

## Downloading a fresh copy from Figshare

The data can be re-ingested at any time to reproduce results against the
original source file. **This requires a network connection and is not
run during vignette build.**

```{r download, eval = FALSE}
raw_path <- download_qibla_data()          # verifies MD5
raw_df   <- import_qibla_data(raw_path)    # reads the XLSX
validate_qibla_data(raw_df)                # structural checks
clean_df <- prepare_qibla_data(raw_df)     # standardise and document
```

The returned object from `prepare_qibla_data()` is identical in structure to
the shipped `gibson_qibla`, with the `provenance` and `audit_log` attributes
populated from the live download.

## Quick descriptive overview

```{r overview}
dplyr::count(gibson_qibla, age_group, sort = TRUE)
```

```{r overview-country}
dplyr::count(gibson_qibla, country, sort = TRUE) |> head(10)
```

## Citing the data

When reporting analyses, cite both this package and the original dataset:

> Gibson, D. (2021). *Early Islamic Qibla Database*.
> figshare. Dataset. <https://doi.org/10.6084/m9.figshare.13570655.v2>
> CC BY 4.0.
