---
title: "Destination-Blind Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Destination-Blind Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## The destination-blind principle

Circular clustering in `qiblalab` is *destination-blind*: cluster assignments
are derived entirely from the observed azimuth data without reference to
Gibson's classification or any candidate destination. This is enforced by
`cluster_qiblas()` (ADR-07).

The workflow is:

1. Cluster orientations into directional groups (unsupervised).
2. Validate the cluster solution internally (compactness, stability).
3. Compare cluster assignments to Gibson's classification *post-hoc* as an
   external validation step — clearly labelled, never used as input.

## Circular summary

Before clustering it is useful to examine the overall directional distribution.

```{r load}
library(qiblalab)

data(gibson_qibla)

# Use mosques with a measured azimuth
mosques <- gibson_qibla[!is.na(gibson_qibla$azimuth), ]
cat(nrow(mosques), "mosques with measured azimuths\n")

circular_summary(mosques$azimuth)
```

The mean resultant length (`rbar`) near 1 indicates tight concentration; near 0
indicates uniformity. The Rayleigh *p*-value tests whether the distribution
departs from uniformity.

## Clustering with circular k-means

```{r cluster}
set.seed(42)
res3 <- cluster_qiblas(gibson_qibla, k = 3, seed = 42)
print(res3)
```

### Model selection

The model selection table shows the total within-cluster dispersion for each
*k* from 1 to `k_max`. Look for an elbow in the curve.

```{r model-sel}
res3$model_selection
```

## Validating the cluster solution

`validate_clusters()` runs three mandatory checks (ADR-10):

1. **Internal**: within-cluster dispersion and centroid separation.
2. **Stability**: bootstrap adjusted Rand index across 50 resamples.
3. **External** (post-hoc): comparison to `gibson_classification` and
   `age_group` — these columns are *not* used to form the clusters.

```{r validate}
val <- validate_clusters(
  res3,
  external      = c("gibson_classification", "age_group"),
  bootstrap_n   = 50L,
  seed          = 1L
)
print(val)
```

### Interpreting the external section

A moderate ARI or NMI against `gibson_classification` is expected if the
orientations carry signal about historical traditions. However, the
destination-blind clusters are *not* a proxy for destination correctness:
two mosques in the same cluster may be oriented toward different destinations
within measurement uncertainty. The external section is evidence for
**directional coherence within clusters**, not destination truth.

## Hierarchical clustering

For comparison, fit the same *k* = 3 with Ward hierarchical clustering.

```{r hierarchical}
res_hc <- cluster_qiblas(gibson_qibla, k = 3,
                          method = "hierarchical", seed = 42)
print(res_hc)
```

Compare the two solutions with the Adjusted Rand Index:

```{r compare-ari}
kmeans_labels <- res3$assignments
hc_labels     <- res_hc$assignments

ok  <- !is.na(kmeans_labels) & !is.na(hc_labels)
ari <- qiblalab:::.ari(kmeans_labels[ok], hc_labels[ok])
cat(sprintf("ARI between k-means and hierarchical solutions: %.3f\n", ari))
```

ARI near 1 means the two methods agree; lower values indicate the
labelling is sensitive to the clustering method, which should be reported.

## Visualising cluster profiles

```{r profiles, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot_cluster_profile(res3)
```

```{r map, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
map_clusters(res3)
```

## Reporting guidance

When publishing results from destination-blind clustering:

- State the method, *k*, seed, and `k_max` scanned.
- Report both internal and stability metrics.
- Label all comparisons with Gibson's classification as *post-hoc external
  validation*, never as model inputs.
- If mean bootstrap ARI < 0.6, report the instability and interpret cluster
  centroids with caution.
- Use structured uncertainty where date ranges are wide; consider re-running
  with `sample_qibla_dataset()` to assess sensitivity.
